Fail Faster

### 2025-03-21 paper/paper.tex
Nit
minor eval edits

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we demostrate that entropy source is an undervalued consideration in generator library design
by isolating its performance contribution, and showing that significant
speedups are possible---even while maintaining output equivalence.
These insights culminate in \name, a recipe for building generator libraries that marry high-level design
with low-level performance.

@@

To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
In both languages, our libraries provide identical functionality and behavior: given the same random seed, a generator written with our technique produces exactly the same sequence of values as an equivalent generator written with the original library.
This semantic equivalence enables pointwise, rather than distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name generators run faster than their unstaged equivalents.
Further, we show generators that run faster also find bugs faster
% ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }
by running these case studies in the Etna
framework~\cite{etna}\bcp{``Etna framework'' sounds like more than
  what it is}, a
platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
think more about how we should talk about Etna} by injecting bugs into a system and
measuring how quickly various generators detect them.

In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions. \jwc{needs a bit about RNG --- BCP thoughts?}
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance:
    abstraction overhead and \rand choice.
    \item We present \name{}, a technique based on staging that eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
    \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements to generator running time. Finally, we show
    that these performance improvements extend to significantly improved bug-finding speed. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding speed'' isn't a race against time.}
    \jwc{We precisely quantify the performance effects of staging and faster rng across a range of testing scenarios --- i.e. a contrib is the fact that we can exactly pin down when one helps versus the other. The ``logically smarter'' your generator is, the more you get out of staging, the ``dumber'', the more you get out of rng.}
    \bcp{Agree with this. We need to beef up our claims here about what we learned about PBT-specifics.}
    \jwc{And this needs to be flowed through the rest of the paper.}
\end{enumerate}

@@

\section{Evaluation}
\label{section:eval}
% \jwc{Initial section here about eval}\tr{todo}
To assess the impact of our approach, we evaluate the performance and bug-finding speed of
\name generators across a range of benchmarks. Our experiments compare generators
built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with
existing generator libraries (and their default randomness mechanisms). Specifically,
we implement
two fully semantically
equivalent staged generator libraries that replicate the behavior of \bq
in OCaml and ScalaCheck in Scala, allowing us to assess the effectiveness of our
technique across different languages and runtime
environments.\bcp{This says ``assess the impact'' twice...}

This section presents our experimental setup, as well as answers to the following research questions:

\begin{itemize}
  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding speed?
  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
\end{itemize}

All experiments were run on a 64-bit Linux machine with 264 GB RAM and a 128-core Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
\scalaname{} and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.

\subsection{Benchmarking Generator Speed}
\label{subsection:benchmarking}
To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to identical ones in \bq,
and generators in \scalaname to those in ScalaCheck.
 We vary the choice of \rand in our \camlname generators, using both \bq's default
\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
in \secref{subsection:faster-rng}.

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a well-typed generator of
STLC terms, whose behavior is the most complex.
For evaluating \scalaname, we implement a subset of the aformentioned
strategies---the boolean list generator, single-pass BST generator, and
well-typed STLC term generator.

The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}.
Our \camlname type-derived BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
dramatically to $6.13-9.65\times$ when combined with \csplitmix. The insertion-based BST generator sees
$1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
using \csplitmix. The single-pass BST generator benefits more significantly from
staging, with speedups of $2.31-2.34\times$, which rise to $8.61-8.95\times$ when combined with
\csplitmix. We attribute the differential in performance speedup to the fact that
BST strategies rely on relatively few calls to generator combinator functions,
but sample extensively from the \rand; notably, single-pass BST, which uses the largest
number of calls to generator combinators, also sees the greatest benefit from staging.

For STLC, staging accounts for the bulk of performance gains, yielding
 $1.71-2.64\times$ speedups for the type-derived STLC generator, and $4.60-5.46\times$ for
 the well-typed generator. When adding \csplitmix, these numbers increase to
 $2.60-5.15\times$ and $4.83-7.75\times$, respectively. This result makes sense because
 STLC generators, particularly the well-typed generator, sample very little from
 randomness libraries, but extensively on generator library functions.

The boolean list generator experiences $3.61-7.10\times$ speedups with
staging, and its performance actually slightly \textit{decreases} when
combined with \csplitmix ($3.43-6.83\times$). This is likely because sampling
booleans is ``cheap'' enough that the overhead of crossing the FFI barrier is
greater than that of generating values directly in OCaml.

@@

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
gain---purely from staging---over ScalaCheck than \camlname does over \bq. This discrepancy arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging allows us to eliminate this
overhead, resulting in substantial speedups in aggregate.

These results show that staging and \rand choice serve distinct but
complementary roles in generator performance. Sampling-heavy generators see
greater improvements from faster randomness libraries, while combinator-heavy
ones benefit more from staging. Since both factors influence performance in
different ways, generator libraries should use both in order to handle diverse
workloads.

@@

\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a varying-size value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better. BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \secref{subsection:faster-rng}'s \csplitmix.}
  \label{fig:time}
\end{figure}

@@

Similarly orthogonal to our approach is techniques for quickly filtering out
inputs that are invalid for testing. Some of these approaches work statically,
by automatically deriving generators that produce valid inputs by
construction~\cite{lampropoulosGeneratingGoodGenerators2017}, while others
filter dynamically via laziness~\cite{claessenGeneratingConstrainedRandom2015}
or by solving satisfiability
problems~\cite{seidelTypeTargetedTesting2015,steinhofelInputInvariants2022,lampropoulosBeginnerLuckLanguage2017}.
In all cases, these approaches could be further improved by insights from \name.

>>>>

We lay a path forward.
First, we show how to use \textit{multi-stage programming}, a classic technique for 
building fast DSLs “without regret”~\cn{}, to eliminate the overhead of 
many common generator abstractions. We refer to this approach as \name.
Second, we show that the source of entropy is an underappreciated factor in 
generator library design by isolating its run-time cost and demonstrating that 
substantial speedups are possible, while maintaining output equivalence.
These insights give a recipe for building generator libraries that marry high-level design
with low-level performance.

@@

To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using ScalaCheck or \bq, respectively.
This semantic equivalence enables pointwise, as opposed to distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name generators run faster than their unstaged equivalents.
Further, we show generators that run faster also find bugs faster
% ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }
by running these case studies in the Etna
platform~\cite{etna}, which simulates real-world PBT usage \iflater\hg{Don't love this phrase, I'll
think more about how we should talk about Etna}\fi by injecting bugs into a system and
measuring how quickly various generators detect them.

In summary, we show that PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions. \jwc{needs a bit about RNG --- BCP thoughts?}
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance:
    abstraction overhead and \rand choice.
    \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
    \item We demonstrate that programs written using \name{} yield performance improvements to generator running time. Finally, we show
    that these performance improvements extend to significantly improved bug-finding speed. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding speed'' isn't a race against time.}
    \jwc{We precisely quantify the performance effects of staging and faster rng across a range of testing scenarios --- i.e. a contrib is the fact that we can exactly pin down when one helps versus the other. The ``logically smarter'' your generator is, the more you get out of staging, the ``dumber'', the more you get out of rng.}
    \bcp{Agree with this. We need to beef up our claims here about what we learned about PBT-specifics.}
    \jwc{And this needs to be flowed through the rest of the paper.}
\end{enumerate}

@@

\section{Evaluation}
\label{section:eval}
% \jwc{Initial section here about eval}\tr{todo}
To assess the impact of our interventions, we evaluate the raw generator performance and bug-finding speed of
\name generators across a range of benchmarks. Our experiments compare generators
built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with
existing generator libraries and their default randomness mechanisms. Specifically,
we implement semantically
equivalent staged generator libraries that replicate the behavior of \bq
in OCaml and ScalaCheck in Scala, allowing us to assess the effectiveness of our
technique across different languages and runtime
environments.\bcp{This says ``assess the impact'' twice...}

This section presents our experimental setup, and address the following research questions:

\begin{itemize}
  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
  \item \textbf{RQ2}: Do observed generation speedups translate to better bug-finding speed?
  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
\end{itemize}

All experiments were run on a 64-bit Linux machine with 264 GB RAM and a 128-core Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
The baseline OCaml generators were written with \bq{} 0.16.
\scalaname{} and all Scala benchmarks
were run on Scala 3.6.3
and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for performance microbenchmarking. \tr{Etna?}

\subsection{Benchmarking Generator Speed}
\label{subsection:benchmarking}
To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to semantically identical ones in \bq,
and generators in \scalaname to those in ScalaCheck.
 We vary the choice of \rand in our \camlname generators, using both \bq's default
\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
in \secref{subsection:faster-rng}.

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
variety of \textit{strategies}, varying in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a generator of STLC terms that is correct by construction (``well-typed''), whose behavior is the most complex.
For evaluating \scalaname, we implement a subset of the aformentioned
strategies---the boolean list generator, single-pass BST generator, and
well-typed STLC term generator.

The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}.
Our \camlname type-derived BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
dramatically, to $6.13-9.65\times$, when combined with \csplitmix. The insertion-based BST generator sees
$1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
using \csplitmix. The single-pass BST generator benefits more significantly from
staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with
\csplitmix. We attribute the difference in performance speedup to the fact that
BST strategies rely on relatively few calls to generator combinator functions,
but sample extensively from the \rand; notably, single-pass BST, which uses the largest
number of calls to generator combinators, also sees the greatest benefit from staging. \tr{Call this out earlier? Unless we have the fancy plot.}

For STLC, staging accounts for the bulk of performance gains, yielding
 $1.71-2.64\times$ speedups for the type-derived STLC generator and $4.60-5.46\times$ for
 the well-typed generator. Adding \csplitmix, these numbers increase to
 $2.60-5.15\times$ and $4.83-7.75\times$. This result makes sense because
 STLC generators, particularly the well-typed generator, sample very little from
 randomness libraries but rely extensively on generator library functions.

The boolean list generator experiences $3.61-7.10\times$ speedups with
staging, and its performance actually slightly \textit{decreases} when
combined with \csplitmix ($3.43-6.83\times$). This is likely because sampling
booleans is cheap enough that the overhead of crossing the FFI barrier is
greater than that of generating values directly in OCaml.

@@

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
gain---purely from staging---over ScalaCheck than \camlname does over \bq. This difference arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging eliminates this
overhead.

These results show that staging and \rand choice play distinct and
complementary roles in generator performance. Sampling-heavy generators see
greater improvements from faster randomness libraries, while combinator-heavy
ones benefit more from staging. Since both factors influence performance 
significantly, generator libraries should use both in order to handle diverse
workloads.

@@

\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate using each strategy. Performance benefits scale with generation size (lower lines translate to greater performance benefit). BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \secref{subsection:faster-rng}'s \csplitmix.}
  \label{fig:time}
\end{figure}

@@

Similarly orthogonal to our approach are techniques for quickly filtering out
inputs that are invalid for testing. Some of these approaches work statically,
by automatically deriving generators that produce valid inputs by
construction~\cite{lampropoulosGeneratingGoodGenerators2017}, while others
filter dynamically via laziness~\cite{claessenGeneratingConstrainedRandom2015}
or by solving satisfiability
problems~\cite{seidelTypeTargetedTesting2015,steinhofelInputInvariants2022,lampropoulosBeginnerLuckLanguage2017}.
In all cases, these approaches could be further improved by insights from \name.
